Perturbing SU($n$)-symmetric interactions

### 2020-01-30 collective_sun/sun_perturbation.tex
modified some derivations in the appendix on operator products

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we use the permutational symmetry of the
ground-state manifold to simplify
\begin{align}
  H_1^{(1)} = \sum_{X,p} v_{Xp} \P_0 X_p \P_0
  \EQFS \sum_{X,p} v_{Xp} X_0
  \EQFS \sum_X \bar v_X \col{X},
  \label{eq:H_1_1}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, and
\begin{align}
  \bar v_X \equiv \f1N \sum_p v_{Xp},
  &&
  \col{X} \equiv \sum_p X_p,
\end{align}
are respectively the mean value in $\v v_X$ and the collective version
of $X$.  In order to compute the second order effective Hamiltonian
$H_1^{(2)}$, we first choose an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \sum_X \sum_p \p{\sum_q h_{pq} v_{Xq} - h_p v_{Xp}} X_p
  \ket\psi,
  &&
  h_p \equiv \sum_q h_{pq},
  \label{eq:H_V1_psi}
\end{align}
where we define $h_{pq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $h_{pq}=h_{qp}$ and $h_{pp}=0$.  We thus find that a
perturbation $\V_1^\Delta\ket\psi$ (indexed by $\Delta$) is an
eigenvector of the interaction Hamiltonian $H_0$ if its coefficient
vectors $\v v_X^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \p{\m h - \diag\v h}\c\v v = \Delta \v v,
  &&
  \m h \equiv \sum_{p,q} h_{pq} \op{p}{q},
  &&
  \v h \equiv \sum_q h_q \ket{q},
  \label{eq:cond_1}
\end{align}
where $\m h$ is a matrix of all $h_{pq}$, and
$\diag\v h\equiv\sum_p h_p \op{p}$ is a matrix with $\v h$ on the
diagonal and zeros everywhere else.  Solving the general eigenvalue
problem in \eqref{eq:cond_1} requires diagonalizing the $N\times N$
matrix $\m h-\diag\v h$.  When the interaction Hamiltonian $H_0$ is
translationally invariant, however, this eigenvalue problem can be
solved analytically; we provide the corresponding solution in Appendix
\ref{sec:trans_inv}.

@@

A classification of terms in \eqref{eq:sym_prod_vec} by index
assignments $g\in\G\p{\m M}$ allows us to expand
\begin{align}
  \prod_{j\in\ZZ_p} \sum_{k\in\ZZ_N^{M_j}} w_j\p{k} O_j\p{k}
  \EQFS \sum_{g\in\G\p{\m M}} \S\p{g} w\p{g} O\p{g},
  \label{eq:sym_prod_group_start}
\end{align}
where $O\p{g}$ is an operator acquired by assigning indices to the
tensors $O_j$ in a manner consistent with $g$; $w\p{g}$ is a scalar
acquired by summing over all indices assigned to the tensors $w_j$
according to $g$; and $\S\p{g}$ is a symmetry factor accounting for
the number of equivalent ways to assign indices according to $g$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Single-body product}
\label{sec:PXYP}

Calculating the second-order effective Hamiltonian $H_1^{(2)}$ induced
on the fully symmetric manifold $\M_0$ by the perturbation $\V_1$ in
\eqref{eq:perturbations} requires us to simplify
\begin{align}
  \P_0 \p{\V_1^\Delta}^2 \P_0
  \EQFS \sum_{\substack{X,Y\\p,q}} v_{Xp} v_{Yq} X_p Y_q
  = \sum_{X,Y,p} v_{Xp} v_{Yp} X_p Y_p
  + \sum_{\substack{X,Y\\p\ne q}} v_{Xp} v_{Yq} X_p Y_q,
  \label{eq:PXYP_start}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$; and all coefficient vectors $\v v_X$
satisfy \eqref{eq:cond_1} with eigenvalue $\Delta\ne0$, which implies
that all $\v v_X$ are mean-zero.  The first sum on the right of
\eqref{eq:PXYP_start} is then
\begin{align}
  \sum_{X,Y,p} v_{Xp} v_{Yp} X_p Y_p
  \EQFS \sum_{X,Y,p} v_{Xp} v_{Yp} X_0 Y_0
  \EQFS \f1N \sum_{X,Y} \v v_X \c\v v_Y \col{XY},
  &&
  \col{Z} \equiv \sum_p Z_p.
  \label{eq:PXYP_eq}
\end{align}
The second sum on the right of \eqref{eq:PXYP_start}, meanwhile, is
\begin{align}
  \sum_{\substack{X,Y\\p\ne q}} v_{Xp} v_{Yq} X_p Y_q
  \EQFS \sum_{\substack{X,Y\\p\ne q}} v_{Xp} v_{Yq} X_0 Y_1,
  \label{eq:PXYP_neq_start}
\end{align}
where we can use the fact that the vectors $\v v_X,\v v_Y$ are
mean-zero to simplify the sum over $p\ne q$ in
\eqref{eq:PXYP_neq_start} to
\begin{align}
  \sum_{p\ne q} v_{Xp} v_{Yq}
  = \sum_{p,q} v_{Xp} v_{Yq} - \sum_{p=q} v_{Xp} v_{Yq}
  = - \v v_X\c\v v_Y,
\end{align}
In order to simplify the product of operators in
\eqref{eq:PXYP_neq_start}, we expand
\begin{align}
  \col{X}\,\col{Y} = \sum_{p,q} X_p Y_q
  = \sum_{p=q} X_p Y_q + \sum_{p\ne q} X_p Y_q
  \EQFS \col{XY} + N\p{N-1} X_0 Y_1,
  \label{eq:PXYP_neq_ops}
\end{align}
which implies that
\begin{align}
  \sum_{\substack{X,Y\\p\ne q}} v_{Xp} v_{Yq} X_p Y_q
  \EQFS - \sum_{X\ne Y} \v v_X\c\v v_Y X_0 Y_1
  \EQFS - \f1{N\p{N-1}}\sum_{X,Y} \v v_X\c\v v_Y
  \p{\col{X}\,\col{Y} - \col{XY}}.
  \label{eq:PXYP_neq}
\end{align}
Altogether, we have that
\begin{align}
  \P_0 \p{\V_1^\Delta}^2 \P_0
  = - \f1{N\p{N-1}} \sum_{X,Y} \v v_X \c\v v_Y
  \p{\col{X}\,\col{Y} - N \col{XY}}.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Two-body product}
\label{sec:POQP}

Calculating the second-order effective Hamiltonian $H_2^{(2)}$ induced
on the fully symmetric manifold $\M_0$ by the perturbation $\V_2$ in
\eqref{eq:perturbations} requires us to simplify
\begin{align}
  \P_0 \p{\V_2^\Delta}^2 \P_0
  = \sum_{O,Q} \sum_{\substack{k<\ell\\p<q}}
  w_{Ok\ell} w_{Qpq} O_{k\ell} Q_{pq}
  \EQFS \f14 \sum_{O,Q} \sum_{k,\ell,p,q}
  w_{Ok\ell} w_{Qpq} O_{k\ell} Q_{pq},
  \label{eq:POQP_start}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$; we define $w_{Opq}$ for all $p,q$ (as
opposed to only $p<q$) by enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$;
and the coefficient vectors $\v w_O$ satisfy the eigenvalue equation
\eqref{eq:cond_2} with eigenvalue $\Delta\ne0$, which implies that all
$\v{\m w}_O$ are mean-zero.  To simplify the sum in
\eqref{eq:POQP_start}, we first consider the terms with
$\p{p,q}=\p{k,\ell}$:
\begin{align}
  \sum_{k<\ell}
  w_{Ok\ell} w_{Qk\ell} O_{k\ell} Q_{k\ell}
  = \v{\m w}_O \c \v{\m w}_Q O_{1,2} Q_{1,2}.
\end{align}
We then consider terms $p\in\set{k,\ell}$ and $q\notin\set{k,\ell}$ or
vice versa:
\begin{align}
  \f14 \sum_{\substack{k,\ell\\p\in\set{k,\ell}\\q\notin\set{k,\ell}}}
  w_{Ok\ell} w_{Qpq} O_{k\ell} Q_{pq}
  &\EQFS \f12 \sum_{\substack{k,\ell\\q\notin\set{k,\ell}}}
  w_{Ok\ell} w_{Qkq} O_{1,2} Q_{2,3} \\
  \f14 \sum_{\substack{k,\ell\\p\notin\set{k,\ell}\\q\in\set{k,\ell}}}
  w_{Ok\ell} w_{Qpq} O_{k\ell} Q_{pq}
  &\EQFS \f12 \sum_{\substack{k,\ell\\q\notin\set{k,\ell}}}
  w_{Ok\ell} w_{Qkq} O_{1,2} Q_{2,3},
\end{align}
where
\begin{align}
  \sum_{\substack{k,\ell\\q\notin\set{k,\ell}}} w_{Ok\ell} w_{Qkq}
  = \sum_{k,\ell,q} w_{Ok\ell} w_{Qkq}
  - \sum_{k,\ell} w_{Ok\ell} w_{Qk\ell}
  = \v w_O\c\v w_Q - 2 \v{\m w}_O \c \v{\m w}_Q.
\end{align}
Finally, the terms in \eqref{eq:POQP_start} with
$p,q\notin\set{k,\ell}$ are
\begin{align}
  \f14 \sum_{\substack{k,\ell\\p,q\notin\set{k,\ell}}}
  w_{Ok\ell} w_{Qpq} O_{k\ell} Q_{pq}
  \EQFS \f14 \sum_{\substack{k,\ell\\p,q\notin\set{k,\ell}}}
  w_{Ok\ell} w_{Qpq} O_{1,2} Q_{3,4},
\end{align}
where, using the fact that all $\v{\m w}_O$ are mean-zero,
\begin{multline}
  \f14 \sum_{\substack{k,\ell\\p,q\notin\set{k,\ell}}} w_{Ok\ell} w_{Qpq}
  = - \f14 \sum_{\substack{k,\ell\\p,q\in\set{k,\ell}}} w_{Ok\ell} w_{Qpq}
  - \f14 \sum_{\substack{k,\ell\\p\in\set{k,\ell}\\q\notin\set{k,\ell}}}
  w_{Ok\ell} w_{Qpq}
  - \f14 \sum_{\substack{k,\ell\\p\notin\set{k,\ell}\\q\in\set{k,\ell}}}
  w_{Ok\ell} w_{Qpq} \\
  = -\v w_O \c \v w_Q + \v{\m w}_O \c \v{\m w}_Q.
\end{multline}
Altogether, we thus have that
\begin{multline}
  \P_0 \p{\V_2^\Delta}^2 \P_0
  \EQFS \sum_{O,Q}
  \bigg[ \v{\m w}_O \c \v{\m w}_Q O_{1,2} Q_{1,2}
  + \p{\v w_O\c\v w_Q - 2 \v{\m w}_O \c \v{\m w}_Q}
  O_{1,2} Q_{2,3} \\
    - \p{\v w_O \c \v w_Q - \v{\m w}_O \c \v{\m w}_Q}
    O_{1,2} Q_{3,4} \bigg].
  \label{eq:POQP}
\end{multline}
In principle, our next task is to write this result in terms of
collective operators within the fully symmetric manifold; this task
turns out to be rather non-trivial in general.  We therefore discuss
the decomposition of multi-local operators (e.g.~$O_{1,2} Q_{2,3}$, a
3-local operator) into a sum of collective operators within the fully
symmetric manifold in Appendix \ref{sec:multi_to_collective}.  At the
end of Appendix \ref{sec:multi_to_collective}, we provide analytical
expressions for the special cases of
\begin{enumerate*}
\item multi-local operators that are tensor powers of a single
  operator, i.e.~of the form $X^{\otimes k}$; and
\item multi-local operators of the form $X^{\otimes k}$ with $X^2=1$.
\end{enumerate*}
These cases are relevant for e.g.~the SU(2) case with Ising-type
interactions of the form $Z\otimes Z$, with $Z$ a single-spin Pauli
operator.

>>>>

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we use the permutational symmetry of the
ground-state manifold to simplify
\begin{align}
  H_1^{(1)} = \sum_{X,p} v_{Xp} \P_0 X_p \P_0
  \EQFS \sum_{X,p} v_{Xp} X_1
  \EQFS \sum_X \bar v_X \col{X},
  \label{eq:H_1_1}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, and
\begin{align}
  \bar v_X \equiv \f1N \sum_p v_{Xp},
  &&
  \col{X} \equiv \sum_p X_p,
\end{align}
are respectively the mean value in $\v v_X$ and the collective version
of $X$.  In order to compute the second order effective Hamiltonian
$H_1^{(2)}$, we first choose an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \sum_X \sum_p \p{\sum_q h_{pq} v_{Xq} - h_p v_{Xp}} X_p
  \ket\psi,
  &&
  h_p \equiv \sum_q h_{pq},
  \label{eq:H_V1_psi}
\end{align}
where we define $h_{pq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $h_{pq}=h_{qp}$ and $h_{pp}=0$.  We thus find that a
perturbation $\V_1^\Delta\ket\psi$ (indexed by $\Delta$) is an
eigenvector of the interaction Hamiltonian $H_0$ if its coefficient
vectors $\v v_X^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \p{\m h - \diag\v h}\c\v v = \Delta \v v,
  &&
  \m h \equiv \sum_{p,q} h_{pq} \op{p}{q},
  &&
  \v h \equiv \sum_q h_q \ket{q},
  \label{eq:cond_1}
\end{align}
where $\m h$ is a matrix of all $h_{pq}$, and
$\diag\v h\equiv\sum_p h_p \op{p}$ is a matrix with $\v h$ on the
diagonal and zeros everywhere else.  Solving the general eigenvalue
problem in \eqref{eq:cond_1} requires diagonalizing the $N\times N$
matrix $\m h-\diag\v h$.  When the interaction Hamiltonian $H_0$ is
translationally invariant, however, this eigenvalue problem can be
solved analytically; we provide the corresponding solution in Appendix
\ref{sec:trans_inv}.

@@

A classification of terms in \eqref{eq:sym_prod_vec} by index
assignments $g\in\G\p{\m M}$ allows us to expand
\begin{align}
  \prod_{j\in\ZZ_p} \sum_{k\in\ZZ_N^{M_j}} w_j\p{k} O_j\p{k}
  \EQFS \sum_{g\in\G\p{\m M}} \S\p{g} w\p{g} O\p{g},
  \label{eq:sym_prod_group_start}
\end{align}
where $\EQFS$ denotes equality up to a restriction to the fully
symmetric manifold, $O\p{g}$ is an operator acquired by assigning
indices to the tensors $O_j$ in a manner consistent with $g$; $w\p{g}$
is a scalar acquired by summing over all indices assigned to the
tensors $w_j$ according to $g$; and $\S\p{g}$ is a symmetry factor
accounting for the number of equivalent ways to assign indices
according to $g$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Product of two single-body operators}
\label{sec:PXYP}

Calculating the second-order effective Hamiltonian $H_1^{(2)}$ induced
on the fully symmetric manifold $\M_0$ by the perturbation $\V_1$ in
\eqref{eq:perturbations} requires us to simplify a product of the form
\begin{align}
  \P_0 \sp{\sum_{p,q} v_{Xp} v_{Yq} X_p Y_q} \P_0
  \EQFS \sum_p v_{Xp} v_{Yp} X_1 Y_1
  + \sum_{p\ne q} v_{Xp} v_{Yq} X_1 Y_2,
  \label{eq:PXYP_start}
\end{align}
where $X,Y$ are single-body operators; and the coefficients
$v_{Xp},v_{Yq}$ satisfy \eqref{eq:cond_1} with eigenvalue
$\Delta\ne0$, which implies that they are mean-zero.  The first sum on
the right of \eqref{eq:PXYP_start} is then
\begin{align}
  \sum_p v_{Xp} v_{Yp} X_1 Y_1
  \EQFS \f1N \v v_X \c \v v_Y\, \col{XY},
  &&
  \col{Z} \equiv \sum_p Z_p.
\end{align}
To simplify the second sum on the right of \eqref{eq:PXYP_start}, we
use the fact that $\v v$ is mean-zero to write
\begin{align}
  \sum_{p\ne q} v_{Xp} v_{Yq}
  = \sum_{p,q} v_{Xp} v_{Yq} - \sum_{p=q} v_{Xp} v_{Yq}
  = - \v v_X \c\v v_Y,
\end{align}
and we expand
\begin{align}
  \col{X}\,\col{Y}
  = \sum_{p=q} X_p Y_q + \sum_{p\ne q} X_p Y_q
  \EQFS \col{XY} + N\p{N-1} X_1 Y_2,
\end{align}
which implies that
\begin{align}
  \sum_{p\ne q} v_{Xp} v_{Yq} X_1 Y_2
  \EQFS - \v v_X\c\v v_Y \p{\col{X}\,\col{Y} - \col{XY}}.
\end{align}
Altogether, we have that
\begin{align}
  \sum_{p,q} v_{Xp} v_{Yq} X_p Y_q
  \EQFS - \f{\v v_X \c\v v_Y}{N\p{N-1}}
  \p{\col{X}\,\col{Y} - N \col{XY}}.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Product of two two-body operators}
\label{sec:POQP}

Calculating the second-order effective Hamiltonian $H_2^{(2)}$ induced
on the fully symmetric manifold $\M_0$ by the perturbation $\V_2$ in
\eqref{eq:perturbations} requires us to simplify
\begin{multline}
  \P_0 \sp{\sum_{\substack{k<l\\p<q}}
    w_{Ok\ell} w_{Qpq} O_{k\ell} Q_{pq}} \P_0
  \EQFS \sum_{p<q} w_{Opq} w_{Qpq} O_{1,2} Q_{1,2}
  + \sum_{\substack{k,p,q\\k\ne p\ne q\ne k}}
  w_{Okp} w_{Qkq} O_{1,2} Q_{1,3} \\
  + \sum_{\substack{k<\ell\\p<q\\\set{k,\ell}\ne\set{p,q}}}
  w_{Ok\ell} w_{Qpq} O_{1,2} Q_{3,4},
  \label{eq:POQP_start}
\end{multline}
where $O$ and $Q$ are permutationally-symmetric two-body operators;
and the coefficients $w_{Ok\ell},w_{Qpq}$ satisfy \eqref{eq:cond_2}
with eigenvalue $\Delta\ne0$, which implies that they are mean-zero.
Defining
\begin{align}
  \v{\m m} \equiv \sum_{p<q} m_{pq} \ket{\set{p,q}},
  &&
  \v m \equiv \sum_p m_p \ket{p},
  &&
  m_p \equiv \sum_q m_{pq},
\end{align}
we can write the first sum on the right of \eqref{eq:POQP_start} as
\begin{align}
  \sum_{p<q} w_{Opq} w_{Qpq} = \v{\m w}_O \c\v{\m w}_Q,
\end{align}
and, using the fact that $w_{Opp}=0$, we can write the second sum on
the right of \eqref{eq:POQP_start} as
\begin{align}
  \sum_{\substack{k,p,q\\k\ne p\ne q\ne k}} w_{Okp} w_{Qkq}
  = \sum_{k,p} w_{Okp} \sum_{q\ne p} w_{Okq}
  = \sum_{k,p} w_{Okp} \p{w_{Qk} - w_{Qkp}}
  = \v w_O \c \v w_Q - 2 \v{\m w}_O \c \v{\m w}_Q.
\end{align}
Using additionally the fact that $w_{Opq},w_{Qpq}$ are mean-zero, we
can similarly simplify
\begin{align}
  \sum_{\substack{p<q\\k<\ell\\\set{k,\ell}\ne\set{p,q}}}
  w_{Ok\ell} w_{Qpq} = -\v w_O \c \v w_Q + \v{\m w}_O \c \v{\m w}_Q.
\end{align}
Altogether, we thus have that
\begin{multline}
  \sum_{\substack{k<l\\p<q}} w_{Ok\ell} w_{Qpq} O_{k\ell} Q_{pq}
  \EQFS \v{\m w}_O \c \v{\m w}_Q O_{1,2} Q_{1,2}
  + \p{\v w_O\c\v w_Q - 2 \v{\m w}_O \c \v{\m w}_Q} O_{1,2} Q_{1,3} \\
  - \p{\v w_O \c \v w_Q - \v{\m w}_O \c \v{\m w}_Q} O_{1,2} Q_{3,4}.
  \label{eq:POQP}
\end{multline}
In principle, our next task is to write this result in terms of
collective operators within the fully symmetric manifold; this task
turns out to be rather non-trivial in general.  We therefore discuss
the decomposition of multi-local operators (e.g.~$O_{1,2} Q_{2,3}$, a
3-local operator) into a sum of collective operators within the fully
symmetric manifold in Appendix \ref{sec:multi_to_collective}.  At the
end of Appendix \ref{sec:multi_to_collective}, we provide analytical
expressions for the special cases of
\begin{enumerate*}
\item multi-local operators that are tensor powers of a single
  operator, i.e.~of the form $X^{\otimes k}$; and
\item multi-local operators of the form $X^{\otimes k}$ with $X^2=1$.
\end{enumerate*}
These cases are relevant for e.g.~the SU(2) case with Ising-type
interactions of the form $Z\otimes Z$, with $Z$ a single-spin Pauli
operator.
